To be FAIR: Theory Specification Needs an Update

### 2025-06-07 action_letter.Rmd
update 2025-06-07 14:19:27.889095

```{asis}
This is a very nice section; I really enjoyed reading this more practical angle!
```

@@

@@

@@

```{asis}
Editors generally prefer that reviewers not preface their reviews with declarations of their lack of expertise in the relevant subjects, but I am going to allow myself to do so here, because I already said as much to the editor when I was asked to review the paper. I have no particular expertise in meta-science; I have never (I am somewhat embarrassed to admit) so much as pre-registered an analysis. I am 71, and open science procedures are the blinking 12:00 on my VCR.

@@

In this case, I think the FAIR framework would be very helpful. The S-R hypothesis makes a number of empirical claims with respect to theory. These claims have changed in response to empirical results and theoretical disputes over the years.  I have often been frustrated by the difficulty of maintaining meaningful boundaries on the theory and the predictions it makes.
```

```{asis}
Here are two examples. In the months after our paper was published, someone (Nagoshi, C. T., & Johnson, R. C. (2005). Socioeconomic status does not moderate the familiality of cognitive abilities in the Hawaii Family Study of Cognition. Journal of Biosocial Science, 37(6), 773-781.) shot out a "failure to replicate." But it wasn't-- their analysis studied parent-child correlations, not twins, and the S-R hypothesis makes no predictions about them. But despite 20 years of trying to make this clear in the published literature, the study is still cited. An even worse example happened recently, when a group of "race scientists" (Pesta, B. J., Kirkegaard, E. O., te Nijenhuis, J., Lasker, J., & Fuerst, J. G. (2020). Racial and ethnic group differences in the heritability of intelligence: A systematic review and meta-analysis. Intelligence, 78, 101408.) used the S-R hypothesis as a platform for a bunch of unrelated claims about race differences.
```

```{asis}
To be clear, it cuts both ways. It has turned out over the years that the S-R interaction does not occur in Europe, and I have revised the theory to fit the new facts, but I have done so on a strictly ad hoc basis. It would be easy to suggest that I am adding epicycles to it, and having a formalized structure for the theoretic revisions would help keep me honest.  Similarly, when conducting new analyses, I am often confronted with investigator degrees of freedom (should we use parental education or income as an SES indicator) that are very difficult to resolve in the privacy of the lab.  Having a structured platform for doing so would help. (This sounds like pre-registration. I don't think there is a bright line between FAIR theorizing and preregistering.)
```

Discuss relationship betwene FAIR theory and prereg? That sounds like a job for Aaron!

@@

>>>>

@@

Done.

@@

We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, up from ~7% of articles in 2000–2009 to ~50% of articles in 2010-2019.
Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:

`r get_revision("datasharing")`

@@

```{asis}
I submit this review signed as Eric Turkheimer.

Editors generally prefer that reviewers not preface their reviews with declarations of their lack of expertise in the relevant subjects, but I am going to allow myself to do so here, because I already said as much to the editor when I was asked to review the paper. I have no particular expertise in meta-science; I have never (I am somewhat embarrassed to admit) so much as pre-registered an analysis. I am 71, and open science procedures are the blinking 12:00 on my VCR.

@@

In this case, I think the FAIR framework would be very helpful. The S-R hypothesis makes a number of empirical claims with respect to theory. These claims have changed in response to empirical results and theoretical disputes over the years.  I have often been frustrated by the difficulty of maintaining meaningful boundaries on the theory and the predictions it makes.

Here are two examples. In the months after our paper was published, someone (Nagoshi, C. T., & Johnson, R. C. (2005). Socioeconomic status does not moderate the familiality of cognitive abilities in the Hawaii Family Study of Cognition. Journal of Biosocial Science, 37(6), 773-781.) shot out a "failure to replicate." But it wasn't-- their analysis studied parent-child correlations, not twins, and the S-R hypothesis makes no predictions about them. But despite 20 years of trying to make this clear in the published literature, the study is still cited. An even worse example happened recently, when a group of "race scientists" (Pesta, B. J., Kirkegaard, E. O., te Nijenhuis, J., Lasker, J., & Fuerst, J. G. (2020). Racial and ethnic group differences in the heritability of intelligence: A systematic review and meta-analysis. Intelligence, 78, 101408.) used the S-R hypothesis as a platform for a bunch of unrelated claims about race differences.

To be clear, it cuts both ways. It has turned out over the years that the S-R interaction does not occur in Europe, and I have revised the theory to fit the new facts, but I have done so on a strictly ad hoc basis. It would be easy to suggest that I am adding epicycles to it, and having a formalized structure for the theoretic revisions would help keep me honest.  Similarly, when conducting new analyses, I am often confronted with investigator degrees of freedom (should we use parental education or income as an SES indicator) that are very difficult to resolve in the privacy of the lab.  Having a structured platform for doing so would help.
```

Dear prof. Turkheimer,

on behalf of all co-authors, I would like to express our sincere appreciation for your thoughtful comments. This is my (Caspar's) favorite review to date, and I especially value your perspective as a seasoned scholar/theoretician. All authors unanimously agreed that it would be a pity to deprive our readers of an opportunity to view FAIR theory through the lens of your lived experience - but we were not sure how to incorporate it. One author proposed to invite you as a co-author, others were concerned that this might constitute a conflict of interest.

We took the initiative to draft a paragraph that summarizes your comments on FAIR theory from the review, and mention your contribution in the Acknowledgements. We would be equally happy to credit your contribution with a co-authorship, depending on your - and the Editor's - preferences.

Our draft paragraph now reads:

`r get_revision("turkheimer")`

The Acknowledgement now reads:

> We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.

```{asis}
(This sounds like pre-registration. I don't think there is a bright line between FAIR theorizing and preregistering.)
```

We think that this comment signals a need to better explain the important distinctions between these two practices.
To address this comment, we have included the following section in the revision:

`r get_revision("fairtheoryprereg")`

@@

`r get_revision("modularpublisingdetract")`

### 2025-06-07 manuscript.Rmd
update 2025-06-07 14:19:27.889095

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Conceptualization
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University department of Methodology & Statistics, Tilburg, The Netherlands"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, Amsterdam, The Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, München, Germany"
  - id            : "7"
    institution   : "MSB Medical School Berlin Department of Psychology, Berlin, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
In most fields, the primary unit of scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable *information artifacts*,
with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the [Crossref](https://www.crossref.org/) association,
which is used by many academic publishers.
Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property `resourceType: dataset`,
and cross-reference the published paper with `relationType: IsSupplementTo`.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`. Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
If an existing theory is updated, `relationType:IsNewVersionOf` could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
<!-- FS: I still think we should have a dedicated term for a digital FAIR theory object -->

@@

To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs)c).

@@

FAIR theory requires a departure from contemporary practice.
Several factors can expedite such a culture change.
One key factor is the *recognition and rewards* movement:
practices for evaluating scientific output are evolving, with initiatives like the [*Declaration on Research Assessment* (DORA)](https://sfdora.org/read/) and [Coalition for Advancing Research Assessment](https://coara.eu/)
promoting the use of more diverse and meaningful metrics beyond journal impact factors.
Modular publishing capitalizes on these changing metrics,
and publishing theories as citeable artifacts allows scholars to be credited for contributions to theory [@kirczModularityNextForm1998].
Journals that publish theoretical papers could require authors to additionally publish their theories in a FAIR format, cross-referencing the paper,
to expedite its effective reuse and iterative improvement.
A second factor is to lower barriers to adopting FAIR theory by building upon existing widely adopted open science infrastructures.
For this reason, we advocate the use of Git for version control, Zenodo for archiving, and DataCite for standardized metadata.
Barriers to entry can also be lowered by simplifying workflows, which is the goal of the `theorytools` R-package.
Fourth, the availability of Open Educational Materials (OEM) about theory development contributes to doctoral socialization.
These materials allow teachers to incorporate theory development into their curriculum without investing substantial time on course development,
thus educating the next generation to make use of and contribute to FAIR theory.
Finally, community building plays an important role;
the international network of open science communities, reproducibility networks, and other similar initiatives provide platforms for disseminating FAIR theories and related methodological innovations.
Authors can also share their FAIR theories with other early adopters by submitting them to the "FAIR Theory Community" on Zenodo.

@@

<span id="fairtheoryprereg">One important future direction is embedding FAIR theories withing existing open science methodologies. 
For example, consider how FAIR theory relates to preregistration.
These practices are clearly distinct but complimentary.
The purpose of FAIR theory is to communicate general principles and expectations about a given phenomenon,
and to provide infrastructure for explicitly deriving hypotheses from specific theories and revising those theories in light of empirical results.
The purpose of preregistration, by contrast, is to eliminate inductive bias from hypothesis tests and increase trust in their outcomes [@peikertWhyDoesPreregistration2023].
These practices differ in their level of abstraction:
FAIR theories cut across studies on a given phenomenon,
whereas preregistrations are specific implementations of hypothesis tests,
with a specific study design, analysis plan, and - optionally - a fully reproducible analysis pipeline. 
These practices complement each other:
authors can make the derivation chain from theory to hypothesis more explicit by citing a specific FAIR theory in their preregistration.
Moreover, it is possible to preregister an inferential procedure that would require revising the theory after observing data,
or even to have proponents and detractors of a theory review a registered report of such a test.
In short, combining FAIR theory with preregistration and other existing open science practices has the potential to strengthen the epistemic cycle of prediction, testing, and revision, moving us closer to a cumulative science.</span>

@@

FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
It provides much-needed open science methods for the inductive phase of the empirical cycle,
closing a critical gap in the scientific process.
FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
This paves the way for more theory-driven scholarship
and accelerates cumulative knowledge acquisition in psychology, the social sciences, and beyond.

>>>>

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Conceptualization
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University department of Methodology & Statistics, Tilburg, The Netherlands"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, Amsterdam, The Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, München, Germany"
  - id            : "7"
    institution   : "MSB Medical School Berlin Department of Psychology, Berlin, Germany"
authornote: |
  We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion. We also thank the Editor, Dr. Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments.
  This version of the paper was generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
The primary unit of scientific communication has long been the academic paper.
In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
These resources are often merely described in papers (if at all), and not made available for reuse.
Modular publishing is the practice of making each of these resources available as independent *information artifacts*,
facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].
<span id="datasharing">Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology [@tedersooDataSharingPractices2021].</span>

Modular publishing can be achieved by archiving specific resources (including theory) in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts.
<!-- - as does, for example, the [Crossref](https://www.crossref.org/) association, -->
<!-- which is used by many academic publishers. -->
The DataCite Metadata Schema offers a standard way to document information artifacts with relevant metadata which increases their findability and documents their relationships to other resources [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, data can be archived in Zenodo with the metadata property `resourceType: dataset`.
If the data were collected for a specific paper, that relationship can be cross-referenced with `relationType: IsSupplementTo`.
Similarly, a FAIR theory can be connected to a specific paper by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`.
Other cross-references are useful for documenting relationships between multiple theory objects: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
If an existing theory is updated, `relationType:IsNewVersionOf` could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.

<span id="modularpublisingdetract">Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication.
</span>

@@

To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs)c).

@@

FAIR theory requires a departure from contemporary practice.
Several factors can expedite such a culture change.
One key factor is the *recognition and rewards* movement:
practices for evaluating scientific output are evolving, with initiatives like the [*Declaration on Research Assessment* (DORA)](https://sfdora.org/read/) and [Coalition for Advancing Research Assessment](https://coara.eu/)
promoting the use of more diverse and meaningful metrics beyond journal impact factors.
Modular publishing capitalizes on these changing metrics,
and FAIR theory allows scholars to be credited for theoretical contributions [@kirczModularityNextForm1998].
Journals that publish theoretical papers could require authors to additionally publish their theories in a FAIR format, cross-referencing the paper,
to expedite its effective reuse and iterative improvement.
A second factor is to lower barriers to adopting FAIR theory by building upon existing widely adopted open science infrastructures.
For this reason, we advocate the use of Git for version control, Zenodo for archiving, and DataCite for standardized metadata.
Barriers to entry can also be lowered by simplifying workflows, which is the goal of the `theorytools` R-package.
Fourth, the availability of Open Educational Materials (OEM) about theory development contributes to doctoral socialization.
These materials allow teachers to incorporate theory development into their curriculum without investing substantial time on course development,
thus educating the next generation to make use of and contribute to FAIR theory.
Finally, community building plays an important role;
the international network of open science communities, reproducibility networks, and other similar initiatives provide platforms for disseminating FAIR theories and related methodological innovations.
Authors can also share their FAIR theories with other early adopters by submitting them to the "FAIR Theory Community" on Zenodo.

@@

<span id="fairtheoryprereg">One important future direction is embedding FAIR theories withing existing open science methodologies. 
For example, consider how FAIR theory relates to preregistration.
These practices are distinct but complimentary.
The purpose of FAIR theory is to communicate general principles and expectations about a given phenomenon,
and to provide infrastructure for explicitly deriving hypotheses from specific theories and revising those theories in light of empirical results.
The purpose of preregistration, by contrast, is to eliminate inductive bias from hypothesis tests and increase trust in the outcomes of a specific empirical study [@peikertWhyDoesPreregistration2023].
FAIR theories are specified at a level of abstraction that transcends individual studies.
FAIR theories can inform - and be informed by - both quantitative and qualitative research.
Preregistrations, by contrast, are specific implementations of quantitative hypothesis tests,
within the context of a specific study design, analysis plan, and - optionally - a fully reproducible analysis pipeline.
These practices complement each other:
authors can make the derivation chain from theory to hypothesis more explicit by citing a specific FAIR theory in their preregistration.
Moreover, it is possible to preregister an inferential procedure that would require revising the theory after observing data,
or even to have proponents and detractors of a theory review a registered report of such a test.
In short, combining FAIR theory with preregistration and other existing open science practices has the potential to strengthen the epistemic cycle of prediction, testing, and revision, moving us closer to a cumulative science.</span>

@@

FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
It provides much-needed open science methods for the inductive phase of the empirical cycle,
closing a critical gap in the scientific process.

<span id = "turkheimer">We invite the reader to ask the rhetorical question raised by our reviewer, prof. Turkheimer, who - reflecting on a long career as a practitioner of theory in psychology - considered "how adopting a FAIR framework would change, improve, or inhibit what I do".
We paraphrase his answer here^[Edited for brevity.], because his lived experience complements our youthful "techno-enthusiasm".
Prof. Turkheimer remarked that a FAIR framework would have been very helpful in maintaining meaningful boundaries between a theory and the predictions it makes,
in tracking how empirical findings result in changes to empirical claims made by a theory,
and in resolving theoretical disputes.
*"FAIR theory provides a structured, machine- and human-readable framework that distills and formalizes the core components of a theory.
Everything published about a theory can continue to exist in its original form,
while a FAIR representation supplements that literature by explicitly outlining the theory’s assumptions, logical structure, and testable implications.
Importantly, FAIR theory can evolve semi-independently from individual papers, offering a persistent and collaborative object that others can reuse, cite, and refine. 
Contributing to the development of a FAIR theory thus becomes a new way of contributing to the theoretical literature - much like writing a high-quality review article, but with the added benefit of interoperability, versioning, and reusability."*</span>

We envision a future where application of the FAIR principles makes theories more *useful*, living up to Herbert Simon's ideal,
enabling scholars to incorporate theory into their workflows in a tanglible way,
providing explicit derivation chains for hypotheses,
applying transparent rules to select the right control variables for causal inference,
and proposing specific changes to existing theories based on empirical results.
This paves the way for more theory-driven scholarship
and accelerates cumulative knowledge acquisition in psychology, the social sciences, and beyond.
